---
title: "Essay 1 - Culture media growth fitness"
output:
  html_document:
    df_print: paged
    number_sections: true
    keep_md: yes
    theme: cerulean
    toc: yes
---


```{r set-options, echo=FALSE, cache=FALSE, include=FALSE}

options(width = 1100)

```


```{r message=FALSE, echo=FALSE, warning=FALSE, include=FALSE}

source("../packages.R")

```

# Exploratory data analysis

## Get raw data

```{r echo=FALSE, warning=FALSE, message=FALSE, rows.print=25}

dataset <- read_excel("../data/aggressiveness-essay.xlsx", sheet = "aggressiveness")
dataset$code = paste0(dataset$species, " | ",  dataset$strain)
dataset$inocPosByEssay = paste0(dataset$essay, " | ",  dataset$inocPos)
dataset$inocPosByClone = paste0(dataset$clone, " | ",  dataset$inocPos)
dataset

```

## Plot data profile

```{r echo=FALSE, message=FALSE, warning=FALSE}

histogram1 <- ggplot(dataset, aes(average)) +
  geom_histogram(aes(y = ..density..), bins = 20) +
  labs(x = "Average mycelial growth (cm)", y = "Density") +
  geom_density(col = 'red')

ggsave(
  plot = histogram1,
  filename = "editable-figures/Histogram_data.svg",
  width = 8
)

histogram1

```

## Remove Control from raw dataset

```{r echo=FALSE, rows.print=25}

(dataset <- dataset %>%
   filter(strain != "Control", !is.na(average)))

```

## Plot observed data as a scatterplot

```{r echo=FALSE, fig.width=18, fig.height=24, message=FALSE, warning=FALSE}

scatter1 = dataset %>%
  
  # Reorder
  mutate(code = fct_reorder(code, desc(code))) %>%

  # Generate a plot
  ggplot(aes(x = day, y = average, color = clone, group = clone, shape = inocPosByEssay)) +
  
    # Include base line for comparisons
    geom_hline(yintercept = 75, linetype = "dashed", color = "grey50") +
    geom_vline(xintercept = 3, linetype = "dashed", color = "grey50") +
    
    # Define graph theme
    theme_bw() +
    theme() + 
    
    # Set point geometry
    geom_point(size = 3) +
  
    # Create a smooth line
    geom_smooth(se = FALSE) +
    
    # Wrap graph by isolate
    facet_wrap(~ code + replicate, ncol = 4) +
    
    # Labels
    labs(x = "Growth time (days)", y = "Culture average diameter (mm)")


ggsave(
  plot = scatter1,
  filename = "editable-figures/Scatter_raw_data.svg",
  width = 18,
  height = 24
)

scatter1

```

## Clear dataset

Remove values with sd close to zero.

```{r echo=FALSE, message=FALSE, warning=FALSE, rows.print=25, fig.height=8}

dataset_clean <- dataset %>%
  group_by(essay, code, clone, inocPos, replicate) %>%
  filter(strain != 2718, sd(average, na.rm = TRUE) > 0, n() == 5)

(dataset_clean %>%
  group_by(code, clone, essay, inocPos, replicate) %>%
  summarise(
    n = n(),
    mean = mean(average),
    sd = sd(average),
    cv = (sd(average) / mean(average)) * 100,
    lowerq = quantile(average)[2],
    upperq = quantile(average)[4],
    iqr = upperq - lowerq
  ))

```

```{r echo=FALSE, message=FALSE, warning=FALSE, rows.print=25, fig.height=8}

boxplot1 <- ggplot(
  dataset_clean %>%
  group_by(code, clone, essay, inocPos, replicate) %>%
  summarise(
    n = n(),
    mean = mean(average),
    sd = sd(average),
    cv = (sd(average) / mean(average)) * 100,
    lowerq = quantile(average)[2],
    upperq = quantile(average)[4],
    iqr = upperq - lowerq
  ),
  aes(x = interaction(code, essay), y = cv)
) + 
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1))

ggsave(
  plot = boxplot1,
  filename = "editable-figures/Boxplot_coefficient_of_variation_clean_data.svg",
  width = 14,
  height = 6
)

boxplot1

```


## Plot observed data as a scatterplot

```{r echo=FALSE, fig.width=18, fig.height=24, message=FALSE, warning=FALSE}

scatter2 = dataset_clean %>%
  
  # Reorder
  mutate(code = fct_reorder(code, desc(code))) %>%

  # Generate a plot
  ggplot(aes(x = day, y = average, color = inocPosByClone, group = inocPosByClone, shape = inocPosByEssay)) +
  
    # Include base line for comparisons
    geom_hline(yintercept = 75, linetype = "dashed", color = "grey50") +
    geom_vline(xintercept = 3, linetype = "dashed", color = "grey50") +
    
    # Define graph theme
    theme_bw() +
    theme() + 
    
    # Set point geometry
    geom_point(size = 3) +
  
    # Create a smooth line
    geom_smooth(se = FALSE) +
    
    # Wrap graph by isolate
    facet_wrap(~ code + replicate, ncol = 4) +
    
    # Labels
    labs(x = "Growth time (days)", y = "Culture average diameter (mm)")


ggsave(
  plot = scatter2,
  filename = "editable-figures/Scatter_observed_clean_data.svg",
  width = 18,
  height = 24
)

scatter2

```

# Comprobatory data analysis

## Test if essay have some effect on essay

```{r echo=FALSE}

lmer_h0_essay <- lmer(
  average ~ day : code : inocPosByClone + (1 |  essay), 
  data = dataset_clean, 
  REML = TRUE
)

summary(lmer_h0_essay)

```

### How much the essay is important to explain about the total variance?

```{r echo=FALSE}

estimates1 <- as.data.frame(VarCorr(lmer_h0_essay))

essay_variance = estimates1[1, "vcov"]
residual_variance1 = estimates1[2, "vcov"]

paste0(
  "Variance: ", 
  round(essay_variance / (essay_variance + residual_variance1) * 100, 10), " %"
)

```

## Test if replicate have some effect on essay

```{r echo=FALSE}

lmer_h0_replicate <- lmer(
  average ~ day : code : inocPosByClone + (1 |  replicate), 
  data = dataset_clean, 
  REML = TRUE
)

summary(lmer_h0_replicate)

```

### How much the replicate is important to explain about the total variance?

```{r echo=FALSE}

estimates2 <- as.data.frame(VarCorr(lmer_h0_replicate))

replicate_variance = estimates2[1, "vcov"]
residual_variance2 = estimates2[2, "vcov"]

paste0(
  "Variance: ", 
  round(replicate_variance / (replicate_variance + residual_variance2) * 100, 10), " %"
)

```

## Generate full model

```{r echo=FALSE}

# Good explanation of LMM in https://www.statsmodels.org/stable/mixed_linear.html

lmer_h1 <- lmer(
  average ~ day : code : inocPosByClone + (1 | day : code : inocPosByClone), 
  data = dataset_clean, 
  REML = FALSE,
)

summary(lmer_h1)

```

### Plot final model adjust

```{r echo=FALSE}

plot(lmer_h1)

qqnorm(resid(lmer_h1))
qqline(resid(lmer_h1))

```

### Test significance of fixed effects

```{r echo=FALSE}

(aov <- anova(lmer_h1))

```

### Test significance of random effects

```{r echo=FALSE}

lmerTest::ranova(lmer_h1)

```

### Plot a residual histogram

```{r echo=FALSE, message=FALSE, warning=FALSE}

lmer_h1_residuals = data.frame(resid(lmer_h1))
colnames(lmer_h1_residuals) <- "Residuals"

histogram2 <- ggplot(lmer_h1_residuals, aes(.data[["Residuals"]])) +
  geom_histogram(aes(y = ..density..), bins = 20) +
  labs(x = "Model residual", y = "Density") +
  geom_density(col = 'red')

ggsave(
  plot = histogram2,
  filename = "editable-figures/Histogram_residuals_full_model.svg",
  width = 8
)

histogram2

```

### Plot observed vs predicted data as a scatterplot

```{r echo=FALSE, fig.width=18, fig.height=8, message=FALSE, warning=FALSE}

scatter3 = dataset_clean %>%
  
  # Reorder
  mutate(code = fct_reorder(code, desc(code))) %>%

  # Generate a plot
  ggplot(aes(x = day, y = average, color = inocPosByClone, group = inocPosByClone)) +
  
    # Include base line for comparisons
    geom_hline(yintercept = 75, linetype = "dashed", color = "grey50") +
    geom_vline(xintercept = 3, linetype = "dashed", color = "grey50") +
    
    # Define graph theme
    theme_bw() +
    theme() + 
    
    # Set point geometry
    geom_point(size = 3) +
  
    geom_line(aes(y = predict(lmer_h1)), size = 1) + 
    
    # Wrap graph by isolate
    facet_wrap(~ code, ncol = 4) +
    
    # Labels
    labs(x = "Growth time (days)", y = "Culture average diameter (mm)")


ggsave(
  plot = scatter3,
  filename = "editable-figures/Scatter_clean_data_observed_vs_predicted.svg",
  width = 14,
  height = 6
)

scatter3

```

## Perform a second clean to dataset

Remove values from the first essay.

```{r echo=FALSE, message=FALSE, warning=FALSE, rows.print=25}

(dataset_clean2 <- dataset_clean %>%
  filter(essay == "Danilo"))

```

### Plot observed reduced data as a scatterplot

```{r echo=FALSE, fig.width=18, fig.height=8, message=FALSE, warning=FALSE}

scatter4 = dataset_clean2 %>%
  
  # Reorder
  #mutate(code = fct_reorder(code, desc(code))) %>%

  # Generate a plot
  ggplot(aes(x = day, y = average, color = inocPosByClone, group = inocPosByClone)) +
  
    # Include base line for comparisons
    geom_hline(yintercept = 75, linetype = "dashed", color = "grey50") +
    geom_hline(yintercept = 25, linetype = "dashed", color = "grey50") +
    geom_vline(xintercept = 3, linetype = "dashed", color = "grey50") +
    
    # Define graph theme
    theme_bw() +
    theme() + 
    
    # Set point geometry
    geom_point(size = 3) +
  
    # Create a smooth line
    geom_smooth(se = FALSE) +
    
    # Wrap graph by isolate
    facet_wrap(~ code, ncol = 4) +
    
    # Labels
    labs(x = "Growth time (days)", y = "Culture average diameter (mm)")


ggsave(
  plot = scatter4,
  filename = "editable-figures/Scatter_clean_data2_observed.svg",
  width = 14,
  height = 6
)

scatter4

```

## Generate reduced model

```{r echo=FALSE}

# Good explanation of LMM in https://www.statsmodels.org/stable/mixed_linear.html

lmer_h2 <- lmer(
  average ~ day : code : clone + (1 | day : code : clone), 
  data = dataset_clean2, 
  REML = FALSE,
)

summary(lmer_h2)

```

### Plot final model adjust

```{r echo=FALSE}

plot(lmer_h2)

qqnorm(resid(lmer_h2))
qqline(resid(lmer_h2))

```

### Test significance of fixed effects

```{r echo=FALSE}

(aov <- anova(lmer_h2))

```

### Test significance of random effects

```{r echo=FALSE}

lmerTest::ranova(lmer_h2)

```

### Plot a residual histogram

```{r echo=FALSE, message=FALSE, warning=FALSE}

lmer_h2_residuals = data.frame(resid(lmer_h2))
colnames(lmer_h2_residuals) <- "Residuals"

histogram3 <- ggplot(lmer_h2_residuals, aes(.data[["Residuals"]])) +
  geom_histogram(aes(y = ..density..), bins = 20) +
  labs(x = "Model residual", y = "Density") +
  geom_density(col = 'red')

ggsave(
  plot = histogram3,
  filename = "editable-figures/Histogram_residuals_reduced_model.svg",
  width = 8
)

histogram3

```

## Plot observed vs predicted with reduced data as a scatterplot

```{r echo=FALSE, fig.width=18, fig.height=8, message=FALSE, warning=FALSE}

scatter5 = dataset_clean2 %>%
  
  # Reorder
  #mutate(code = fct_reorder(code, desc(code))) %>%

  # Generate a plot
  ggplot(aes(x = day, y = average, color = clone, group = clone)) +
  
    # Include base line for comparisons
    geom_hline(yintercept = 75, linetype = "dashed", color = "grey50") +
    geom_hline(yintercept = 25, linetype = "dashed", color = "grey50") +
    geom_vline(xintercept = 3, linetype = "dashed", color = "grey50") +
    
    # Define graph theme
    theme_bw() +
    theme() + 
    
    # Set point geometry
    geom_point(size = 3) +
    
    # Create a smooth line
    geom_line(aes(y = predict(lmer_h2)), size = 1) + 
    
    # Wrap graph by isolate
    facet_wrap(~ code, ncol = 4) +
    
    # Labels
    labs(x = "Growth time (days)", y = "Culture average diameter (mm)")


ggsave(
  plot = scatter5,
  filename = "editable-figures/Scatter_clean_data2_observed_vs_predicted.svg",
  width = 14,
  height = 6
)

scatter5

```

### Generate pairwise comparisons

```{r echo=FALSE, message=FALSE, warning=FALSE, rows.print=25}

# Generate pairwise comparisons
lmeans_lmer_h2 <- lsmeans(
  lmer_h2,
  pairwise ~ day : code : clone
)

# Print comparisons
data.frame(lmeans_lmer_h2$contrasts)

```

### Include alpha-numeric indicators of comparisons significance

```{r echo=FALSE, rows.print=40, paged.print=FALSE}

multicomp_with_letters <- cld(lmeans_lmer_h2$lsmeans, alpha = 0.01, Letters = c(LETTERS))
multicomp_with_letters

```

### Plot multiplicity comparisons

```{r echo=FALSE, fig.width=12, fig.height=8}

plot(
  multicomp_with_letters, 
  type = "response"
)

```

### Plot estimates

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=14, fig.height=6}

stringSanitizer <- function(x) {
  for (item in 1:dim(x)[1]) {
    x[item, ".group"] = gsub(" ", "", x[item, ".group"], fixed = TRUE)
  }

  return(x)
}

comparisons_data <- stringSanitizer(data.frame(multicomp_with_letters))

barplot = comparisons_data %>%
  
  #mutate(cultureMedia = fct_reorder(cultureMedia, lsmean)) %>%
  
  ggplot(aes(x = code, y = lsmean, group = clone, fill = code)) +
    
    # Set error bar
    geom_errorbar(
      aes(ymin = lower.CL, ymax = upper.CL), #, col = species
      width = .1
    ) +
    
    # Set lines
    geom_bar(
      stat = "identity", 
      position = position_dodge(),
      color = "grey30"
      # data = comparisons_data, 
      # color = comparisons_data$species, 
      # group = comparisons_data$species
    ) +
    
    scale_fill_grey() +
    
    # Set theme
    theme_bw() +
    theme(
      legend.position = "none",
      # axis.ticks = element_blank()
    ) +
    
    
    # Grid graph for clone
    facet_grid(. ~ clone) +
    
    # Include base line for comparisons
    geom_hline(yintercept = 60, linetype = "dashed", color = "grey50") +
    geom_hline(yintercept = 20, linetype = "dashed", color = "grey50") +
    
    # Add annotations
    geom_text(
      aes(x = code, y = upper.CL + 5, label = .group),
      fontface = "bold"
    ) +
    
    # Add annotations  
    annotate("text", x = 2.5, y = -5, label = "P. palm.", cex = 4, colour = "grey20") +
    geom_segment(aes(x = 1, xend = 4, y = -2, yend = -2), colour = "grey20") +
    annotate("text", x = 6.5, y = -5, label = "P. theobr. sp. nov.", cex = 4, colour = "grey20") +
    geom_segment(aes(x = 5, xend = 8, y = -2, yend = -2), colour = "grey20") +
    scale_x_discrete(
      #breaks = c(1,2,3,4,5,6,7,8), 
      labels = as.character(c("1102","1158","906","920","1091","1151","1205","1285"))
    ) +
    
    # Set labels
    labs(
      x = expression(paste(italic("Phytophthora"), " strain by Cocoa clone.")),
      y = "Mean growth on the third experiment day (mm)",
      color = "Culture media"
    )

ggsave(
  plot = barplot,
  filename = "editable-figures/Barplot_predicted_reduced_model.svg",
  width = 14,
  height = 6
)

barplot

```
