---
title: "Essay 2 - Temperature growth fitness"
output:
  html_document:
    df_print: paged
    number_sections: true
    keep_md: yes
    theme: cerulean
    toc: yes
---


```{r set-options, echo=FALSE, cache=FALSE, include=FALSE}

options(width = 1200)

```


```{r message=FALSE, echo=FALSE, warning=FALSE, include=FALSE}

source("../packages.R")

```

# Exploratory data analysis

## Get raw data

```{r echo=FALSE, rows.print=25}

dataset <- read_excel("../data/temperature-essay.xlsx", sheet = "temperature")
dataset$code = paste0(dataset$species, " | ",  dataset$strain)
dataset

```

## Plot data profile

```{r echo=FALSE, message=FALSE, warning=FALSE}

histogram1 <- ggplot(dataset, aes(average)) +
  geom_histogram(aes(y = ..density..), bins = 20) +
  labs(x = "Average mycelial growth (cm)", y = "Density") +
  geom_density(col = 'red')

ggsave(
  plot = histogram1,
  filename = "editable-figures/Histogram_data.svg",
  width = 8
)

histogram1

```

## Plot observed data as a boxplot with reduced groups iteraction

```{r echo=FALSE, fig.width=14, fig.height=6, message=FALSE, warning=FALSE}

boxplot1 = dataset %>%
  
  # Reorder
  mutate(species = fct_reorder(species, desc(species))) %>%

  # Generate a plot
  ggplot(aes(x = temperatureCD, y = average, fill = temperatureCD, group = temperatureCD)) +
  
    # Define graph theme
    theme_bw() +
    theme(legend.position = c(0.9, 0.25)) + 
    
    # Set color scale for temperature gradient
    scale_fill_gradient(low = "blue", high = "red") + 
    
    # Set point geometry
    geom_boxplot() +
    
    # Wrap graph by isolate
    facet_wrap(~ code, ncol = 5) +
    
    # Include base line for comparisons
    geom_hline(yintercept = 40, linetype = "dashed", color = "grey50") +
    
    # Labels
    labs(
      x = "Growth temperature ºC", 
      y = "Culture average diameter (mm)",
      color = "Temperature ºC"
    )


ggsave(
  plot = boxplot1,
  filename = "editable-figures/Boxplot_data_reduced.svg",
  width = 8
)

boxplot1

```

## Plot observed data as a boxplot with full groups iteraction

```{r echo=FALSE, fig.width=18, fig.height=18, message=FALSE, warning=FALSE}

boxplot2 = dataset %>%
  
  # Reorder
  mutate(species = fct_reorder(species, desc(species))) %>%

  # Generate a plot
  ggplot(aes(x = temperatureCD, y = average, fill = temperatureCD, group = temperatureCD)) +
  
    # Define graph theme
    theme_bw() +
      
    # Set color scale for temperature gradient
    scale_fill_gradient(low = "blue", high = "red") +
    
    # Set point geometry
    geom_boxplot() +
    
    # Wrap graph by isolate
    facet_wrap(~ code + day, ncol = 7) +
    
    # Include base line for comparisons
    geom_hline(yintercept = 40, linetype = "dashed", color = "grey50") +
    geom_vline(xintercept = 20, linetype = "dashed", color = "grey50") +
    
    # Labels
    labs(
      x = "Growth temperature ºC", 
      y = "Culture average diameter (mm)",
      color = "Temperature ºC"
    )


ggsave(
  plot = boxplot2,
  filename = "editable-figures/Boxplot_data_full.svg",
  width = 8
)

boxplot2

```

## Clear dataset

Remove temperatures 5, 35 and 40.

```{r echo=FALSE, message=FALSE, warning=FALSE, rows.print=15}

dataset_clear <- dataset %>%
   filter(temperatureCD > 5, temperatureCD < 35, !is.na(average))

(dataset_clear <- rbind(
  filter(dataset_clear, essay == "Jennifer"),
  filter(dataset_clear, essay == "Danilo" & temperatureCD > 10)
))

```

## Plot observed data as a scatterplot

```{r echo=FALSE, fig.width=14, fig.height=6, message=FALSE, warning=FALSE}

scatter1 = dataset_clear %>%
  
  # Reorder
  mutate(species = fct_reorder(species, desc(species))) %>%

  # Generate a plot
  ggplot(aes(x = day, y = average, color = temperatureCD, group = temperatureCD)) +
  
    # Include base line for comparisons
    geom_hline(yintercept = 40, linetype = "dashed", color = "grey50") +
    geom_vline(xintercept = 4, linetype = "dashed", color = "grey50") +
    
    # Define graph theme
    theme_bw() +
    theme(legend.position = c(0.9, 0.25)) + 
    
    # Set color scale for temperature gradient
    scale_color_gradient(low = "blue", high = "red") +
    
    # Set point geometry
    geom_point() +
    
    # Wrap graph by isolate
    facet_wrap(~ code, ncol = 5) +
    
    # Labels
    labs(x = "Growth time (days)", y = "Culture average diameter (mm)")


ggsave(
  plot = scatter1,
  filename = "editable-figures/Scatter_raw_data.svg",
  width = 14,
  height = 6
)

scatter1

```

# Comprobatory data analysis

## Test if essay have some effect on essay

```{r echo=FALSE}

lmer_h0_essay <- lmer(
  average ~ day : code : temperatureCategorical + (1 |  essay), 
  data = dataset_clear, 
  REML = TRUE
)

summary(lmer_h0_essay)

```

### How much the essay is important to explain about the total variance?

```{r echo=FALSE}

estimates1 <- as.data.frame(VarCorr(lmer_h0_essay))

essay_variance = estimates1[1, "vcov"]
residual_variance1 = estimates1[2, "vcov"]

paste0(
  "Variance: ", 
  round(essay_variance / (essay_variance + residual_variance1) * 100, 10), " %"
)

```

## Test if replicate have some effect on essay

```{r echo=FALSE}

lmer_h0_replicate <- lmer(
  average ~ day : strain : temperatureCategorical + (1 |  replicate), 
  data = dataset_clear, 
  REML = TRUE
)

summary(lmer_h0_replicate)

```

### How much the replicate is important to explain about the total variance?

```{r echo=FALSE}

estimates2 <- as.data.frame(VarCorr(lmer_h0_replicate))

replicate_variance = estimates2[1, "vcov"]
residual_variance2 = estimates2[2, "vcov"]

paste0(
  "Variance: ", 
  round(replicate_variance / (replicate_variance + residual_variance2) * 100, 10), " %"
)

```

## Generate full model

```{r echo=FALSE}

# Good explanation of LMM in https://www.statsmodels.org/stable/mixed_linear.html

lmer_h1 <- lmer(
  average ~ day : code : temperatureCategorical + (1 | day / strain / temperatureCategorical), 
  data = dataset_clear, 
  REML = FALSE,
)

summary(lmer_h1)

```

### How much the target effects are important to explain about the total variance?

```{r}

estimates3 <- as.data.frame(VarCorr(lmer_h1))

cultureMedia_strain_day_variance = estimates3[1, "vcov"]
strain_day_variance = estimates3[2, "vcov"]
day_variance = estimates3[3, "vcov"]
residual_variance3 = estimates3[4, "vcov"]


# cultureMedia_strain_day_variance
paste0(
  "Variance temperatureCategorical:(strain:day): ", 
  round(cultureMedia_strain_day_variance / (cultureMedia_strain_day_variance + residual_variance3) * 100, 10), " %"
)

# strain_day_variance
paste0(
  "Variance strain:day: ", 
  round(strain_day_variance / (strain_day_variance + residual_variance3) * 100, 10), " %"
)

# day_variance
paste0(
  "Variance day: ", 
  round(day_variance / (day_variance + residual_variance3) * 100, 10), " %"
)

```

### Plot final model adjust

```{r echo=FALSE}

plot(lmer_h1)

qqnorm(resid(lmer_h1))
qqline(resid(lmer_h1))

```


### Test significance of fixed effects

```{r echo=FALSE}

(aov <- anova(lmer_h1))

```

### Test significance of random effects

```{r echo=FALSE}

lmerTest::ranova(lmer_h1)

```

### Plot a residual histogram

```{r echo=FALSE, message=FALSE, warning=FALSE}

lmer_h1_residuals = data.frame(resid(lmer_h1))
colnames(lmer_h1_residuals) <- "Residuals"

histogram2 <- ggplot(lmer_h1_residuals, aes(.data[["Residuals"]])) +
  geom_histogram(aes(y = ..density..), bins = 20) +
  labs(x = "Model residual", y = "Density") +
  geom_density(col = 'red')

ggsave(
  plot = histogram2,
  filename = "editable-figures/Histogram_residuals_full_model.svg",
  width = 8
)

histogram2

```

### Get general adjust of full model

```{r message=FALSE, warning=FALSE, echo=FALSE}

adjust <- cbind(dataset_clear, pred = predict(lmer_h1))

ggplot(adjust, aes(x = average, y = pred)) +
  geom_point() + 
  geom_smooth(method = lm, se = FALSE, fullrange = TRUE) +
  theme_classic() + 
  scale_color_grey() +
  annotate(
    "text", 
    label = paste0("Correlation: ", round(cor(adjust$pred, adjust$average), 3)), 
    x = 10, 
    y = 80
  )

```


### Plot a scatter for observed vs predicted values

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.width=14, fig.height=6}

scatter2 = dataset_clear %>%
  
  # Reorder
  mutate(species = fct_reorder(species, desc(species))) %>%

  # Generate a plot
  ggplot(aes(x = day, y = average, color = temperatureCD, group = temperatureCD)) +
    
    # Horizontal lines
    geom_hline(yintercept = 20, linetype = "dashed", color = "grey70") +
    geom_hline(yintercept = 60, linetype = "dashed", color = "grey70") +
    geom_vline(xintercept = 4, linetype = "dashed", color = "grey70") +
  
    # Define graph theme
    theme_bw() +
    theme(legend.position = c(0.9, 0.25)) + 
    
    # Set color scale for temperature gradient
    scale_color_gradient(low = "blue", high = "red") + 
    
    # Set point geometry
    geom_point() +
    
    # Wrap graph by isolate
    facet_wrap(~ code, ncol = 5) +
    
    # Include predict lines
    geom_smooth(data = cbind(dataset_clear, pred = predict(lmer_h1)), aes(y = pred), size = 1) +
    
    # Labels
    labs(x = "Growth time (days)", y = "Culture average diameter (mm)")


ggsave(
  plot = scatter2,
  filename = "editable-figures/Scatter_observed_and_predicted_full_model.svg",
  width = 14,
  height = 6
)

scatter2

```

### Generate pairwise comparisons

```{r echo=FALSE, message=FALSE, warning=FALSE, rows.print=25}

# Generate pairwise comparisons
lmeans_lmer_h1 <- lsmeans(
  lmer_h1,
  pairwise ~ day : code : temperatureCategorical
)

# Print comparisons
data.frame(lmeans_lmer_h1$contrasts)

```

### Include alpha-numeric indicators of comparisons significance

```{r echo=FALSE, rows.print=40, paged.print=FALSE}

multicomp_with_letters <- cld(lmeans_lmer_h1$lsmeans)
multicomp_with_letters

```

### Plot multiplicity comparisons

```{r echo=FALSE, fig.width=12, fig.height=8}

plot(
  multicomp_with_letters, 
  type = "response"
)

```

## Generate reduced model

```{r echo=FALSE}

lmer_h2 <- lmer(
  average ~ day : species : temperatureCategorical + (1 | day / species / temperatureCategorical), 
  data = dataset_clear, 
  REML = FALSE
)

summary(lmer_h2)

```

### How much the target effects are important to explain about the total variance?

```{r echo=FALSE}

estimates4 <- as.data.frame(VarCorr(lmer_h2))

cultureMedia_strain_day_variance2 = estimates4[1, "vcov"]
strain_day_variance2 = estimates4[2, "vcov"]
day_variance2 = estimates4[3, "vcov"]
residual_variance4 = estimates4[4, "vcov"]


# cultureMedia_strain_day_variance
paste0(
  "Variance temperatureCategorical:(strain:day): ", 
  round(cultureMedia_strain_day_variance2 / (cultureMedia_strain_day_variance2 + residual_variance4) * 100, 10), " %"
)

# strain_day_variance
paste0(
  "Variance strain:day: ", 
  round(strain_day_variance2 / (strain_day_variance2 + residual_variance4) * 100, 10), " %"
)

# day_variance
paste0(
  "Variance day: ", 
  round(day_variance2 / (day_variance2 + residual_variance4) * 100, 10), " %"
)

```

### Plot final model adjust

```{r echo=FALSE}

plot(lmer_h2)

qqnorm(resid(lmer_h2))
qqline(resid(lmer_h2))

```


### Test significance of fixed effects

```{r echo=FALSE}

(aov <- anova(lmer_h2))

```

### Test significance of random effects

```{r echo=FALSE}

lmerTest::ranova(lmer_h2)

```

### Plot a residual histogram

```{r echo=FALSE, message=FALSE, warning=FALSE}

lmer_h2_residuals = data.frame(resid(lmer_h2))
colnames(lmer_h2_residuals) <- "Residuals"

histogram3 <- ggplot(lmer_h2_residuals, aes(.data[["Residuals"]])) +
  geom_histogram(aes(y = ..density..), bins = 20) +
  labs(x = "Model residual", y = "Density") +
  geom_density(col = 'red')

ggsave(
  plot = histogram3,
  filename = "editable-figures/Histogram_residuals_reduced_model.svg",
  width = 8
)

histogram3

```

### Get general adjust of reduced model

```{r message=FALSE, warning=FALSE, echo=FALSE}

adjust <- cbind(dataset_clear, pred = predict(lmer_h2))

adjust2_plot <- ggplot(adjust, aes(x = average, y = pred)) +
  geom_point() + 
  geom_smooth(method = lm, se = FALSE, fullrange = TRUE) +
  theme_classic() + 
  scale_color_grey() +
  annotate(
    "text", 
    label = paste0("Correlation: ", round(cor(adjust$pred, adjust$average), 3)), 
    x = 10, 
    y = 80
  ) +
  
  labs(
    x = "Observed", 
    y = "Predicted"
  )

ggsave(
  plot = adjust2_plot,
  filename = "editable-figures/Adjust_residuals_reduced_model.svg",
  width = 3,
  height = 3
)

adjust2_plot

```

### Generate pairwise comparisons

```{r echo=FALSE, message=FALSE, warning=FALSE, rows.print=25}

# Generate pairwise comparisons
lmeans_lmer_h2 <- lsmeans(
  lmer_h2,
  pairwise ~ day : species : temperatureCategorical
)

# Print comparisons
data.frame(lmeans_lmer_h2$contrasts)

```

### Include alpha-numeric indicators of comparisons significance

```{r echo=FALSE, rows.print=40, paged.print=FALSE}

multicomp_with_letters2 <- cld(lmeans_lmer_h2$lsmeans)
multicomp_with_letters2

```

### Plot multiplicity comparisons

```{r echo=FALSE, fig.width=8, fig.height=4}

plot(
  multicomp_with_letters2, 
  type = "response"
)

```


### Plot a scatter for observed vs predicted values of reduced model

```{r echo=FALSE, fig.height=6, fig.width=8}

stringSanitizer <- function(x) {
  for (item in 1:dim(x)[1]) {
    x[item, ".group"] = gsub(" ", "", x[item, ".group"], fixed = TRUE)
  }

  return(x)
}

comparisons_data <- stringSanitizer(data.frame(multicomp_with_letters2))

scatter3 <- ggplot(comparisons_data, aes(x = temperatureCategorical, y = lsmean)) +
  
  # Set error bar
  geom_errorbar(
    aes(ymin = lower.CL, ymax = upper.CL, col = species), 
    width = .1
  ) +
  
  # Set lines
  geom_line(
    data = comparisons_data, 
    mapping = aes(x = temperatureCategorical, y = lsmean, color = species, group = species), 
    size = 0.5
  ) +
  
  # Set points
  geom_point(
    data = comparisons_data, 
    mapping = aes(x = temperatureCategorical, y = lsmean, color = species), 
    cex = 3
  ) +
  
  # Scale colors
  scale_colour_manual(
    name = "Taxa",
    labels = c("Phytophthora citrophthora", "Phytophthora palmivora", "Phytophthora theobromicola sp. nov."),
    values = c("grey80", "grey60", "black")
  ) +
  
  # Set theme
  theme_bw() +
  
  # Configure aditionals for theme
  theme(
    legend.position = c(.01, .99),
    legend.justification = c("left", "top"),
    legend.box.just = "left",
    legend.box.background = element_rect()
  ) +
  
  # Set custom xaxis labels
  scale_x_discrete(
    breaks = c("T2","T3","T4", "T5", "T6"), 
    labels = c(10, 15, 20, 25, 30)
  ) + 
  
  # Add annotations
  geom_text(
    aes(y = lsmean, label = .group),
    nudge_x = 0.1,
    check_overlap = TRUE,
    fontface = "bold"
  ) + 
  
  # Set labels
  labs(shape = "Taxon") +
  xlab("Temperature (°C)") +
  ylab("Mean growth on the fourth experiment day (mm)")

# ggsave(
#   plot = scatter3,
#   filename = "editable-figures/Scatter_predicted_reduced_model.svg",
#   height = 5,
#   width = 7
# )

scatter3

```


